---
always_allow_html: yes
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
    message=FALSE,
  warning=FALSE
)
# library(rIMPD)
# library(sf)
# library(tidyverse)
# library(lubridate)
# library(leaflet)
# library(htmltools)
# library(mapview)
# impd_meta <- search_impd() %>%
#     st_as_sf(coords = c("longitude", "latitude"), crs=4326)
```

# rIMPD

<!-- badges: start -->

[![R build status](https://github.com/chguiterman/rIMPD/workflows/R-CMD-check/badge.svg)](https://github.com/chguiterman/rIMPD/actions)

<!-- badges: end -->

Explore North American Fire-Scar Records from the International Multiproxy Paleofire Database

`rIMPD` provides basic functions to explore and extract tree-ring data stored on the International Multiproxy Paleofire Database (IMPD), hosted by the National Centers for Environmental Information under the US National Oceanic and Atmospheric Administration. The IMPD falls under the World Data Service for Paleoclimatology (WDS-Paleo): <https://www.ncei.noaa.gov/products/paleoclimatology>, and has become the *defacto* warehouse for publicly-accessible tree-ring fire history data.

For an overview of the package with some useful examples, run the following code: `vignette(topic = "Introduction-to-rIMPD", package = "rIMPD")`

## Usage

Using `rIMPD`, you can discover the fire history sites across the North American continent:

```{r}
library(rIMPD)
library(ggplot2)
library(dplyr)
library(sf)

all_sites <- search_impd() %>% 
    st_as_sf(coords = c("longitude", "latitude"), crs=4326) 

ggplot() +
  geom_sf(data = spData::world) +
  geom_sf(data = all_sites, color = "purple") +
  coord_sf(xlim=c(-165, -50), ylim=c(15, 70)) +
  labs(title = "North American Tree-Ring Records on the IMPD",
       caption = paste0("Data current as of ", lubridate::today())
       ) +
  theme_void()


```

The tree-ring data associated with each site can be obtained directly:

```{r}
site_to_get <- all_sites[10, ]$studyCode
fhx <- get_impd_fhx(site_to_get)
```

and analyzed with tools from the `burnr` library (Malevich et al. 2018), for example:

```{r}
library(burnr)
plot_demograph(fhx, 
               composite_rug = TRUE,
               plot_legend = TRUE) +
  labs(title = all_sites[10, ]$siteName, 
       subtitle = paste0("Collected by: ", all_sites[10, ]$investigators),
       caption = "Tree-ring fire history plot. Each horizontal line represents the life-span
       of a tree, with vertical tick marks denoting  fire scars recorded in the tree rings. 
       At bottom, vertical dashes indicate 'widespread' fires that were recorded on 25% of 
       the trees.")
```

Funding for `rIMPD` was provided by the USGS Community for Data Integration <https://www.usgs.gov/centers/cdi>

## Disclaimer

This software is in the public domain because it contains materials that originally came from the U.S. Geological Survey, an agency of the United States Department of Interior. For more information, see the official USGS copyright policy.

Although this software program has been used by the U.S. Geological Survey (USGS), no warranty, expressed or implied, is made by the USGS or the U.S. Government as to the accuracy and functioning of the program and related program material nor shall the fact of distribution constitute any such warranty, and no responsibility is assumed by the USGS in connection therewith.

This software is provided "AS IS."

## Reference

Malevich, S.B., Guiterman, C.H., and Margolis, E.Q (2018). `burnr`: Fire History Analysis and Graphics in R. Dendrochronologia 49 (June): 9--15. <https://doi.org/10.1016/j.dendro.2018.02.005>
